---
title: "maternal_mortality_case_study"
output:
  pdf_document: default
  html_document: default
date: "2022-10-04"
---

# Initial Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(root.dir = "/home/mireles/horseless/maternal_mortality/mortality")
```

Here, we load necessary libraries:

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```
It's pertinent to verify that we are in the right working directory.

```{r}
getwd()
```
Because the dataset is loaded as characters, there are columns that need to be converted to other datatypes. Many of these are factors instead of characters. There are also some columns that need to be added for the sake of processing later. Because we are combining many years worth of data, that is a variable to keep track of.

```{r}
preprocess_dataset <- function(mort, year=0) {
  mort$res_status <- as_factor(mort$res_status)
  mort$education <- as_factor(mort$education)
  mort$age <- as.numeric(mort$age)
  mort$month <- as.numeric(mort$month)
  mort$death_loc <- as_factor(mort$death_loc)
  mort$marital_status <- as_factor(mort$marital_status)
  mort$race <- as_factor(mort$race)
  #mort$hisp_orig <- as_factor(mort$hisp_orig)
  mort$hisp_orig_race <- as_factor(mort$hisp_orig_race)
  
  mort <- mort %>%
    mutate(age = replace(age, age == 999, NA))
  
  if (year != 0) {
    mort <- mort %>% add_column(year=year, .after="month")
    mort$year <- as_factor(mort$year)
  }
  
  return(mort)
}
```

Loading all data, processing it, and combining it into one variable.

```{r, warning=FALSE, message=FALSE}
mort2019 <- read_csv("maternal_mortality_2019.csv")
mort2019 <- preprocess_dataset(mort2019, 2019)

mort2018 <- read_csv("maternal_mortality_2018.csv")
mort2018 <- preprocess_dataset(mort2018, 2018)

mort2017 <- read_csv("maternal_mortality_2017.csv")
mort2017 <- preprocess_dataset(mort2017, 2017)

mort2016 <- read_csv("maternal_mortality_2016.csv")
mort2016 <- preprocess_dataset(mort2016, 2016)

mort2015 <- read_csv("maternal_mortality_2015.csv")
mort2015 <- preprocess_dataset(mort2015, 2015)

mort2014 <- read_csv("maternal_mortality_2014.csv")
mort2014 <- preprocess_dataset(mort2014, 2014)

mort2013 <- read_csv("maternal_mortality_2013.csv")
mort2013 <- preprocess_dataset(mort2013, 2013)

mort2012 <- read_csv("maternal_mortality_2012.csv")
mort2012 <- preprocess_dataset(mort2012, 2012)

mort2011 <- read_csv("maternal_mortality_2011.csv")
mort2011 <- preprocess_dataset(mort2011, 2011)

mort2010 <- read_csv("maternal_mortality_2010.csv")
mort2010 <- preprocess_dataset(mort2010, 2010)

mort2009 <- read_csv("maternal_mortality_2009.csv")
mort2009 <- preprocess_dataset(mort2009, 2009)

mort2008 <- read_csv("maternal_mortality_2008.csv")
mort2008 <- preprocess_dataset(mort2008, 2008)

mort2007 <- read_csv("maternal_mortality_2007.csv")
mort2007 <- preprocess_dataset(mort2007, 2007)

mort2006 <- read_csv("maternal_mortality_2006.csv")
mort2006 <- preprocess_dataset(mort2006, 2006)

mort2005 <- read_csv("maternal_mortality_2005.csv")
mort2005 <- preprocess_dataset(mort2005, 2005)

mort2004 <- read_csv("maternal_mortality_2004.csv")
mort2004 <- preprocess_dataset(mort2004, 2004)

mort2003 <- read_csv("maternal_mortality_2003.csv")
mort2003 <- preprocess_dataset(mort2003, 2003)

mort_combined <- bind_rows(mort2019, mort2018, mort2017, mort2016,
                          mort2015, mort2014, mort2013, mort2012,
                          mort2011, mort2010, mort2009, mort2008,
                          mort2007, mort2006, mort2005, mort2004,
                          mort2003)
```

There are more housekeeping functions that we need. The following returns a list of all codes associated with a patient, with the option to include the code of the underlying condition.

```{r}
get_codes <- function(row, include_underlying=FALSE) {
  underlying <- which(colnames(row)=="underlying")
  first_code <- which(colnames(row)=="code1")
  last_code <- which(colnames(row)=="code20")

  if (include_underlying == TRUE) {
    codes <- row[c(underlying:last_code)]
  } else {
    codes <- row[c(first_code:last_code)]
  }

  a_codes <- list()
  for (c in codes) {
    if (is.na(c)) {
      return(a_codes)
    } else {
      a_codes <- append(a_codes, c)
    }
  }
  return(a_codes)
}
```

The following takes a mortality tibble as an argument and returns a version that includes a column with the number of codes associated with the patient, not counting the underlying condition.

```{r}
add_num_conditions <- function(mort) {
  num_conditions_col <- c()
  for (row in 1:nrow(mort)) {
    codes <- get_codes(mort[row,], include_underlying=FALSE)
    num_conditions_col <- append(num_conditions_col, length(codes))
  }

  mort <- mort %>% add_column(num_conditions=num_conditions_col, .before="underlying")
  return(mort)
}
```

Interestingly, the ICD-10 groups even the underlying conditions into broader categories that may be useful for our analysis. The following is a set of regular expressions that follows the groupings outlined by the ICD-10 groupings relating to pregnancy, childbirth, and the puerperium at https://www.icd10data.com/ICD10CM/Codes/O00-O9A.

```{r}
code_buckets <- c("O0[0-8]", # pregnancy with abortive outcome
                  "O09",     # Supervision of high-risk pregnancy
                  "O1[0-6]", # Edema, proteinuria, and hypertensive
                  "O2[0-9]", # Other maternal disorders predominantly related to pregnancy
                  "O[3|4][0-8]|O39",   # Maternal care related to the fetus and amniotic cavity and possible delivery problems
                  "O[6][0-9]|O7[0-7]", #  Complications of labor and delivery
                  "O8[0-2]", # Encounter for delivery
                  "O8[5-9]|O9[0-2]", # Complications predominantly related to the puerperium
                  "O9[4-9]|O9A" # Other obstetric conditions, not elsewhere classified
)
```

```{r}
group_codes <- function(mort, code) {
  bucket <- mort[0,]
  for (row in 1:nrow(mort)) {
    underlying <- mort[row, "underlying"]
    if (!is.na(str_extract(underlying, code))) {
      bucket <- bucket %>% add_row(mort[row,])
    }
  }

  return(bucket)
}

```

The tibble is divided into buckets. It was less obvious before because the codes are opaque to the untrained eye, but here we can see that we have omitted code buckets 2 and 7, representing supervision with high-risk pregnancy and encounter for for delivery, respectively.

According to [Outsource Strategies International](https://www.outsourcestrategies.com/resources/high-risk-pregnancy-icd-10-coding-changes-2017/), O09 is only intended for use in the prenatal period. It is therefore not directly applicable to this case study. Encounter for pregnancy codes are meant specifically for situations where there were no complications during delivery, so it is not immediately relevant here.

Both of these buckets are not represented in the data at the moment, but they are worth keeping in mind as other datasets are brought in and the analysis expands.

```{r}
abortive <- group_codes(mort_combined, code_buckets[1])
eph <- group_codes(mort_combined, code_buckets[3])
preg <- group_codes(mort_combined, code_buckets[4])
fetal <- group_codes(mort_combined, code_buckets[5])
comp <- group_codes(mort_combined, code_buckets[6])
puerperium <- group_codes(mort_combined, code_buckets[8])
other <- group_codes(mort_combined, code_buckets[9])

#buckets = lst(abortive, eph, preg, fetal, comp, puerperium, other)
```

Just for verification of the above:

```{r}
# TODO: clean this in the final draft
high_risk <- group_codes(mort_combined, code_buckets[2])
encounter <- group_codes(mort_combined, code_buckets[7])
head(high_risk)
head(encounter)
```

```{r}
add_bucket <- function(mort, bucket_name) {
  mort <- mort %>% add_column(bucket=bucket_name, .before="underlying")
  return(mort)
}
```

Here we combine individual buckets back into a single tibble, this time with a column giving each patient the respective broader bucket their underlying condition falls into.

```{r}
mort <- mort_combined[0,]
abortive <- add_bucket(abortive, "abortive")
eph <- add_bucket(eph, "eph")
comp <- add_bucket(comp, "comp")
preg <- add_bucket(preg, "preg")
fetal <- add_bucket(fetal, "fetal")
puerperium <- add_bucket(puerperium, "puerperium")
other <- add_bucket(other, "other")
```

```{r}
mort <- abortive %>%
  bind_rows(eph, comp, preg, fetal, puerperium, other)
```

Here we add the number of conditions to the main dataset. The underlying condition is excluded because the codes following it always have a more specific version of that code, so it would be redundant.

```{r}
mort <- add_num_conditions(mort)
```

# Making Things Legible
The codes are very densy packed and not especially human readable. Fortunately, it will not be difficult to change the tibble such that anyone can see at least the broad demographic characteristics of a given patient. First, we decode `res_status`, or the resident status. This is perhaps confusingly named, as the codes here are indicative of whether the patient is a resident of the state and county (residents), the state but not the county (intrastate residents), out of state but still from the US, and foreign residents.

```{r}
mort$res_status_decoded <- as.numeric(mort$res_status)
mort <- mort %>%
  filter(!is.na(res_status_decoded)) %>%
  mutate(res_status_decoded = case_when(res_status == 1 ~ "resident",
                                res_status == 2 ~ "intrastate nonresident",
                                res_status == 3 ~ "interstate nonresident",
                                res_status == 4 ~ "foreign resident")) %>%
  relocate(res_status_decoded, .after="res_status")
```

We do the same for place of death.

```{r}
mort$death_loc_decoded <- as.numeric(mort$death_loc)
mort <- mort %>%
  filter(!is.na(death_loc_decoded)) %>%
  mutate(death_loc_decoded = case_when(death_loc == 1 ~ "Inpatient - hospital, clinic, or medical center",
                                       death_loc == 2 ~ "Outpatient or admitted to ER - hospital, clinicl, or medical center",
                                       death_loc == 3 ~ "Dead on arrival - hospital, clinic, or medical center",
                                       death_loc == 4 ~ "Home",
                                       death_loc == 5 ~ "Hospice facility",
                                       death_loc == 6 ~ "Nursing home/long term care",
                                       death_loc == 7 ~ "Other",
                                       death_loc == 9 ~ "Place of death unknown")) %>%
  relocate(death_loc_decoded, .after="death_loc")
```

We do the same for education.

```{r, warning=FALSE}
mort$education_decoded <- as.numeric(mort$education)
mort <- mort %>%
  filter(!is.na(education)) %>%
  mutate(education_decoded = case_when(education == 1 ~ "8th grade or less",
                               education == 2 ~ "9th-12th grade, no diploma",
                               education == 3 ~ "high school graduate or GED completed",
                               education == 4 ~ "some college credit, but no degree",
                               education == 5 ~ "associate's degree",
                               education == 6 ~ "bachelor's degree",
                               education == 7 ~ "master's degree",
                               education == 8 ~ "doctorate or professional degree",
                               education == 9 ~ "unknown")) %>%
  relocate(education_decoded, .after="education")
```

And for race and Hispanic origins.

```{r}
mort$race_decoded <- as.character(mort$race)
mort <- mort %>%
  filter(!is.na(race_decoded)) %>%
  mutate(race_decoded = case_when(race == "01" ~ "White",
                                  race == "02" ~ "Black",
                                  race == "03" ~ "American Indian (includes Aleuts and Eskimos",
                                  race == "04" ~ "Chinese",
                                  race == "05" ~ "Japanese",
                                  race == "06" ~ "Hawaiian (includes Part-Hawaiian",
                                  race == "07" ~ "Filipino",
                                  race == "18" ~ "Asian Indian",
                                  race == "28" ~ "Korean",
                                  race == "38" ~ "Samoan",
                                  race == "48" ~ "Vietnamese",
                                  race == "58" ~ "Guamanian",
                                  race == "68" ~ "Other Asian or Pacific Islander",
                                  race == "78" ~ "Combined other Asian or Pacific Islander")) %>%
  relocate(race_decoded, .after="race")
```

```{r}
mort$hisp_orig_decoded <- mort$hisp_orig
mort <- mort %>%
  filter(!is.na(hisp_orig_decoded)) %>%
  mutate(hisp_orig_decoded = case_when(hisp_orig >= 100 & hisp_orig <= 199 ~ "Non-Hispanic",
                                       hisp_orig >= 200 & hisp_orig <= 209 ~ "Spaniard",
                                       hisp_orig >= 210 & hisp_orig <= 219 ~ "Mexican",
                                       hisp_orig >= 221 & hisp_orig <= 230 ~ "Central American",
                                       hisp_orig >= 231 & hisp_orig <= 249 ~ "South American",
                                       hisp_orig >= 250 & hisp_orig <= 259 ~ "Latin American",
                                       hisp_orig >= 260 & hisp_orig <= 269 ~ "Puerto Rican",
                                       hisp_orig >= 270 & hisp_orig <= 274 ~ "Cuban",
                                       hisp_orig >= 275 & hisp_orig <= 279 ~ "Dominican",
                                       hisp_orig == 220 ~ "Central or South American",
                                       hisp_orig >= 280 & hisp_orig <= 299 ~ "Other Hispanic",
                                       hisp_orig >= 996 & hisp_orig <= 999 ~ "Unknown")) %>%
  relocate(hisp_orig_decoded, .after="hisp_orig")
```

Finally as a bit of housekeeping, there is a recode that broadly combined Hispanic origin and race that we will handle here.

```{r}
mort$hisp_orig_race_decoded <- mort$hisp_orig_race
mort <- mort %>%
  filter(!is.na(hisp_orig_race_decoded)) %>%
  mutate(hisp_orig_race_decoded = case_when(hisp_orig_race == 1 ~ "Mexican",
                                            hisp_orig_race == 2 ~ "Puerto Rican",
                                            hisp_orig_race == 3 ~ "Cuban",
                                            hisp_orig_race == 4 ~ "Central or South American",
                                            hisp_orig_race == 5 ~ "Other or unknown Hispanic",
                                            hisp_orig_race == 6 ~ "Non-Hispanic White",
                                            hisp_orig_race == 7 ~ "Non-Hispanic Black",
                                            hisp_orig_race == 8 ~ "Non-Hispanic Other Races",
                                            hisp_orig_race == 9 ~ "Hispanic origin unknown")) %>%
  relocate(hisp_orig_race_decoded, .after="hisp_orig_race")
```