---
title: "maternal_mortality_case_study"
output:
  pdf_document: default
  html_document: default
date: "2022-10-04"
---

# Initial Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(root.dir = "/home/mireles/horseless/maternal_mortality/mortality")
```

Here, we load necessary libraries:

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```
It's pertinent to verify that we are in the right working directory.

```{r}
getwd()
```
Because the dataset is loaded as characters, there are columns that need to be converted to other datatypes. Many of these are factors instead of characters. There are also some columns that need to be added for the sake of processing later. Because we are combining many years worth of data, that is a variable to keep track of.

```{r}
preprocess_dataset <- function(mort, year=0) {
  mort$res_status <- as_factor(mort$res_status)
  mort$education <- as_factor(mort$education)
  mort$age <- as.numeric(mort$age)
  mort$month <- as.numeric(mort$month)
  mort$death_loc <- as_factor(mort$death_loc)
  mort$marital_status <- as_factor(mort$marital_status)
  mort$race <- as_factor(mort$race)
  #mort$hisp_orig <- as_factor(mort$hisp_orig)
  mort$hisp_orig_race <- as_factor(mort$hisp_orig_race)
  
  mort <- mort %>%
    mutate(age = replace(age, age == 999, NA))
  
  if (year != 0) {
    mort <- mort %>% add_column(year=year, .after="month")
    mort$year <- as.numeric(mort$year)
  }
  
  return(mort)
}
```

Loading all data, processing it, and combining it into one variable.

```{r, warning=FALSE, message=FALSE}
mort2019 <- read_csv("maternal_mortality_2019.csv")
mort2019 <- preprocess_dataset(mort2019, 2019)

mort2018 <- read_csv("maternal_mortality_2018.csv")
mort2018 <- preprocess_dataset(mort2018, 2018)

mort2017 <- read_csv("maternal_mortality_2017.csv")
mort2017 <- preprocess_dataset(mort2017, 2017)

mort2016 <- read_csv("maternal_mortality_2016.csv")
mort2016 <- preprocess_dataset(mort2016, 2016)

mort2015 <- read_csv("maternal_mortality_2015.csv")
mort2015 <- preprocess_dataset(mort2015, 2015)

mort2014 <- read_csv("maternal_mortality_2014.csv")
mort2014 <- preprocess_dataset(mort2014, 2014)

mort2013 <- read_csv("maternal_mortality_2013.csv")
mort2013 <- preprocess_dataset(mort2013, 2013)

mort2012 <- read_csv("maternal_mortality_2012.csv")
mort2012 <- preprocess_dataset(mort2012, 2012)

mort2011 <- read_csv("maternal_mortality_2011.csv")
mort2011 <- preprocess_dataset(mort2011, 2011)

mort2010 <- read_csv("maternal_mortality_2010.csv")
mort2010 <- preprocess_dataset(mort2010, 2010)

mort2009 <- read_csv("maternal_mortality_2009.csv")
mort2009 <- preprocess_dataset(mort2009, 2009)

mort2008 <- read_csv("maternal_mortality_2008.csv")
mort2008 <- preprocess_dataset(mort2008, 2008)

mort2007 <- read_csv("maternal_mortality_2007.csv")
mort2007 <- preprocess_dataset(mort2007, 2007)

mort2006 <- read_csv("maternal_mortality_2006.csv")
mort2006 <- preprocess_dataset(mort2006, 2006)

mort2005 <- read_csv("maternal_mortality_2005.csv")
mort2005 <- preprocess_dataset(mort2005, 2005)

mort2004 <- read_csv("maternal_mortality_2004.csv")
mort2004 <- preprocess_dataset(mort2004, 2004)

mort2003 <- read_csv("maternal_mortality_2003.csv")
mort2003 <- preprocess_dataset(mort2003, 2003)

mort_combined <- bind_rows(mort2019, mort2018, mort2017, mort2016,
                          mort2015, mort2014, mort2013, mort2012,
                          mort2011, mort2010, mort2009, mort2008,
                          mort2007, mort2006, mort2005, mort2004,
                          mort2003)
```

There are more housekeeping functions that we need. The following returns a list of all codes associated with a patient, with the option to include the code of the underlying condition.

```{r}
get_codes <- function(row, include_underlying=FALSE) {
  underlying <- which(colnames(row)=="underlying")
  first_code <- which(colnames(row)=="code1")
  last_code <- which(colnames(row)=="code20")

  if (include_underlying == TRUE) {
    codes <- row[c(underlying:last_code)]
  } else {
    codes <- row[c(first_code:last_code)]
  }

  a_codes <- list()
  for (c in codes) {
    if (is.na(c)) {
      return(a_codes)
    } else {
      a_codes <- append(a_codes, c)
    }
  }
  return(a_codes)
}
```

The following takes a mortality tibble as an argument and returns a version that includes a column with the number of codes associated with the patient, not counting the underlying condition.

```{r}
add_num_conditions <- function(mort) {
  num_conditions_col <- c()
  for (row in 1:nrow(mort)) {
    codes <- get_codes(mort[row,], include_underlying=FALSE)
    num_conditions_col <- append(num_conditions_col, length(codes))
  }

  mort <- mort %>% add_column(num_conditions=num_conditions_col, .before="underlying")
  return(mort)
}
```

Interestingly, the ICD-10 groups even the underlying conditions into broader categories that may be useful for our analysis. The following is a set of regular expressions that follows the groupings outlined by the ICD-10 groupings relating to pregnancy, childbirth, and the puerperium at https://www.icd10data.com/ICD10CM/Codes/O00-O9A.

```{r}
code_buckets <- c("O0[0-8]", # pregnancy with abortive outcome
                  "O09",     # Supervision of high-risk pregnancy
                  "O1[0-6]", # Edema, proteinuria, and hypertensive
                  "O2[0-9]", # Other maternal disorders predominantly related to pregnancy
                  "O[3|4][0-8]|O39",   # Maternal care related to the fetus and amniotic cavity and possible delivery problems
                  "O[6][0-9]|O7[0-7]", #  Complications of labor and delivery
                  "O8[0-2]", # Encounter for delivery
                  "O8[5-9]|O9[0-2]", # Complications predominantly related to the puerperium
                  "O9[4-9]|O9A" # Other obstetric conditions, not elsewhere classified
)
```

```{r}
group_codes <- function(mort, code) {
  bucket <- mort[0,]
  for (row in 1:nrow(mort)) {
    underlying <- mort[row, "underlying"]
    if (!is.na(str_extract(underlying, code))) {
      bucket <- bucket %>% add_row(mort[row,])
    }
  }

  return(bucket)
}
```

The tibble is divided into buckets. It was less obvious before because the codes are opaque to the untrained eye, but here we can see that we have omitted code buckets 2 and 7, representing supervision with high-risk pregnancy and encounter for for delivery, respectively.

According to [Outsource Strategies International](https://www.outsourcestrategies.com/resources/high-risk-pregnancy-icd-10-coding-changes-2017/), O09 is only intended for use in the prenatal period. It is therefore not directly applicable to this case study. Encounter for pregnancy codes are meant specifically for situations where there were no complications during delivery, so it is not immediately relevant here.

Both of these buckets are not represented in the data at the moment, but they are worth keeping in mind as other datasets are brought in and the analysis expands.

```{r}
abortive <- group_codes(mort_combined, code_buckets[1])
eph <- group_codes(mort_combined, code_buckets[3])
preg <- group_codes(mort_combined, code_buckets[4])
fetal <- group_codes(mort_combined, code_buckets[5])
comp <- group_codes(mort_combined, code_buckets[6])
puerperium <- group_codes(mort_combined, code_buckets[8])
other <- group_codes(mort_combined, code_buckets[9])

#buckets = lst(abortive, eph, preg, fetal, comp, puerperium, other)
```

Just for verification of the above:

```{r}
# TODO: clean this in the final draft
high_risk <- group_codes(mort_combined, code_buckets[2])
encounter <- group_codes(mort_combined, code_buckets[7])
head(high_risk)
head(encounter)
```

```{r}
add_bucket <- function(mort, bucket_name) {
  mort <- mort %>% add_column(bucket=bucket_name, .before="underlying")
  return(mort)
}
```

Here we combine individual buckets back into a single tibble, this time with a column giving each patient the respective broader bucket their underlying condition falls into.

```{r}
mort <- mort_combined[0,]
abortive <- add_bucket(abortive, "abortive")
eph <- add_bucket(eph, "eph")
comp <- add_bucket(comp, "comp")
preg <- add_bucket(preg, "preg")
fetal <- add_bucket(fetal, "fetal")
puerperium <- add_bucket(puerperium, "puerperium")
other <- add_bucket(other, "other")
```

```{r}
mort <- mort %>%
  bind_rows(abortive, eph, comp, preg, fetal, puerperium, other)
```

Here we add the number of conditions to the main dataset. The underlying condition is excluded because the codes following it always have a more specific version of that code, so it would be redundant.

```{r}
mort <- add_num_conditions(mort)
```

# Making Things Legible
The codes are very densy packed and not especially human readable. Fortunately, it will not be difficult to change the tibble such that anyone can see at least the broad demographic characteristics of a given patient. First, we decode `res_status`, or the resident status. This is perhaps confusingly named, as the codes here are indicative of whether the patient is a resident of the state and county (residents), the state but not the county (intrastate residents), out of state but still from the US, and foreign residents.

```{r}
mort$res_status_decoded <- as.numeric(mort$res_status)
mort <- mort %>%
  mutate(res_status_decoded = case_when(res_status == 1 ~ "resident",
                                res_status == 2 ~ "intrastate nonresident",
                                res_status == 3 ~ "interstate nonresident",
                                res_status == 4 ~ "foreign resident")) %>%
  relocate(res_status_decoded, .after="res_status")
```

We do the same for place of death.

```{r}
mort$death_loc_decoded <- as.numeric(mort$death_loc)
mort <- mort %>%
  mutate(death_loc_decoded = case_when(death_loc == 1 ~ "Inpatient - hospital, clinic, or medical center",
                                       death_loc == 2 ~ "Outpatient or admitted to ER - hospital, clinicl, or medical center",
                                       death_loc == 3 ~ "Dead on arrival - hospital, clinic, or medical center",
                                       death_loc == 4 ~ "Home",
                                       death_loc == 5 ~ "Hospice facility",
                                       death_loc == 6 ~ "Nursing home/long term care",
                                       death_loc == 7 ~ "Other",
                                       death_loc == 9 ~ "Place of death unknown")) %>%
  relocate(death_loc_decoded, .after="death_loc")
```

We do the same for education.

```{r, warning=FALSE}
mort$education_decoded <- as.numeric(mort$education)
mort <- mort %>%
  mutate(education_decoded = case_when(education == 1 ~ "8th grade or less",
                               education == 2 ~ "9th-12th grade, no diploma",
                               education == 3 ~ "high school graduate or GED completed",
                               education == 4 ~ "some college credit, but no degree",
                               education == 5 ~ "associate's degree",
                               education == 6 ~ "bachelor's degree",
                               education == 7 ~ "master's degree",
                               education == 8 ~ "doctorate or professional degree",
                               education == 9 ~ "unknown")) %>%
  relocate(education_decoded, .after="education")
```

And for race and Hispanic origins.

```{r}
mort$race_decoded <- as.character(mort$race)
mort <- mort %>%
  mutate(race_decoded = case_when(race == "01" ~ "White",
                                  race == "02" ~ "Black",
                                  race == "03" ~ "American Indian (includes Aleuts and Eskimos",
                                  race == "04" ~ "Chinese",
                                  race == "05" ~ "Japanese",
                                  race == "06" ~ "Hawaiian (includes Part-Hawaiian",
                                  race == "07" ~ "Filipino",
                                  race == "18" ~ "Asian Indian",
                                  race == "28" ~ "Korean",
                                  race == "38" ~ "Samoan",
                                  race == "48" ~ "Vietnamese",
                                  race == "58" ~ "Guamanian",
                                  race == "68" ~ "Other Asian or Pacific Islander",
                                  race == "78" ~ "Combined other Asian or Pacific Islander")) %>%
  relocate(race_decoded, .after="race")
```

```{r}
mort$hisp_orig_decoded <- mort$hisp_orig
mort <- mort %>%
  mutate(hisp_orig_decoded = case_when(hisp_orig >= 100 & hisp_orig <= 199 ~ "Non-Hispanic",
                                       hisp_orig >= 200 & hisp_orig <= 209 ~ "Spaniard",
                                       hisp_orig >= 210 & hisp_orig <= 219 ~ "Mexican",
                                       hisp_orig >= 221 & hisp_orig <= 230 ~ "Central American",
                                       hisp_orig >= 231 & hisp_orig <= 249 ~ "South American",
                                       hisp_orig >= 250 & hisp_orig <= 259 ~ "Latin American",
                                       hisp_orig >= 260 & hisp_orig <= 269 ~ "Puerto Rican",
                                       hisp_orig >= 270 & hisp_orig <= 274 ~ "Cuban",
                                       hisp_orig >= 275 & hisp_orig <= 279 ~ "Dominican",
                                       hisp_orig == 220 ~ "Central or South American",
                                       hisp_orig >= 280 & hisp_orig <= 299 ~ "Other Hispanic",
                                       hisp_orig >= 996 & hisp_orig <= 999 ~ "Unknown")) %>%
  relocate(hisp_orig_decoded, .after="hisp_orig")
```

Finally as a bit of housekeeping, there is a recode that broadly combined Hispanic origin and race that we will handle here.

```{r}
mort$hisp_orig_race_decoded <- mort$hisp_orig_race
mort <- mort %>%
  mutate(hisp_orig_race_decoded = case_when(hisp_orig_race == 1 ~ "Mexican",
                                            hisp_orig_race == 2 ~ "Puerto Rican",
                                            hisp_orig_race == 3 ~ "Cuban",
                                            hisp_orig_race == 4 ~ "Central or South American",
                                            hisp_orig_race == 5 ~ "Other or unknown Hispanic",
                                            hisp_orig_race == 6 ~ "Non-Hispanic White",
                                            hisp_orig_race == 7 ~ "Non-Hispanic Black",
                                            hisp_orig_race == 8 ~ "Non-Hispanic Other Races",
                                            hisp_orig_race == 9 ~ "Hispanic origin unknown")) %>%
  relocate(hisp_orig_race_decoded, .after="hisp_orig_race")
```

## Entry to the Exploration
First, we should start looking at some summary statistics to get a sense of where to start.

```{r}
annual_total <- mort %>%
  group_by(year) %>%
  summarise(annual_total = n())
  
ggplot(annual_total, aes(year, annual_total)) +
  geom_line(group=1) +
  geom_point() +
  ylim(327,900) +
  labs(title="Total number of annual maternal deaths, 2003-2019",
       x="Year",
       y="Annual total")
```
This only gives us the total number of annual maternal deaths. To determine proportions for ourselves, we m

```{r, warning=FALSE}
# This only needs to be done once
# TODO: write code that only executes this if the demographic table
# hasn't already been downlodaed
library(rvest)
library(httr)
library(polite)

url <- "https://en.wikipedia.org/wiki/Demographics_of_the_United_States"

url_bow <- polite::bow(url)
url_bow

ind_html <-
  polite::scrape(url_bow) %>%
  rvest::html_nodes("table.wikitable") %>%
  rvest::html_table(fill = TRUE)

pop_table <- ind_html[[10]]
pop_table <- rename(pop_table, year=names(pop_table)[1])
```

```{r}
library(stringr)
pop_table$year <- str_extract_all(pop_table$year, "\\d{4}", simplify=TRUE)

pop_table_tmp <- pop_table[69:85,]
```

```{r}
pop_table_tmp <- rename(pop_table_tmp, average_population="Average population[67][29][30]")
pop_table_tmp <- rename(pop_table_tmp, live_births="Live births[68]")
pop_table_tmp$year <- as.numeric(pop_table_tmp$year)
```

```{r}
avg_pop <- as.numeric(gsub(",","",pop_table_tmp$average_population))
live_births <- as.numeric(gsub(",","",pop_table_tmp$live_births))
pop_table_tmp$average_population <- avg_pop
pop_table_tmp$live_births <- live_births
```

This will need to be move to the end, after all of the processing of this table has been done.

```{r}
write.table(pop_table_tmp, file="demographic_table.csv", sep=",", row.names=FALSE)
```

```{r}
pop_table <- pop_table_tmp
```

```{r}
annual_total <- annual_total
ratio <- annual_total$annual_total/pop_table$live_births * 100000
annual_total$per_100k <- ratio
ggplot(annual_total, aes(x=year, y=per_100k)) +
  geom_line(group=1) +
  geom_point() +
  ylim(10, 25) +
  labs(title="Maternal deaths per 100,000 live births, 2003-2019",
       x="Year",
       y="Maternal deaths per 100,000 live births")
```

Determine if the average number of codes assigned to a patient have changed.

```{r}
avg_num_codes <- mort %>%
  group_by(year) %>%
  summarize(num_patients = n(),
            num_codes = sum(num_conditions),
            avg_codes = num_codes / num_patients)

ggplot(avg_num_codes, aes(x=year, y=avg_codes)) +
         geom_line(group=1) +
         geom_point() +
         ylim(1, 3) +
         labs(title="Average # of codes per patient, 2003-2019",
              x="Year",
              y="Avg. # of codes")
```

Looking at these two, we notice a sharp increase in both the number of patients and the average number of codes that they present with starting in 2018. My first thought is to drill down and see what codes are common to the last two years that were not common before.

The following pipe generates a tibble that lists the number of occurrences of each code in a given year.

```{r}
common_codes <- mort %>%
  group_by(year) %>%
  add_count(underlying)

common_code_counts <- common_codes %>%
  group_by(year) %>%
  select(year, underlying, n) %>%
  unique() %>%
  slice_max(order_by = n, n = 5) %>%
  arrange(desc(year))

View(common_code_counts)
# Determine how many codes are responsible for most of the cases
# (top 5 for each year)
# and what percentage of cases those were
unique_underlying <- unique(common_code_counts$underlying)

# We see that the top five conditions from each year are
# responsible for almost 81% of our data.
underlying_percentage <- mort %>%
  select(underlying) %>%
  filter(underlying %in% unique_underlying) %>%
  summarise(percentage = n())
underlying_percentage <- underlying_percentage / nrow(mort) * 100
underlying_percentage
```

While documenting these other conditions, I noticed that the code O95 cannot be found on the main ICD-10 site.

```{r}
o95 <- mort %>%
  select(year, underlying,) %>%
  filter(underlying == "O95") %>%
  count(year)

ggplot(o95, aes(x=year, y=n)) +
         geom_line(group=1) + 
         geom_point() +
         labs(title="Number of occurrences of O95, 2003-2019",
              x="Year",
              y="Number")
```

Graphing the changes of each of the top coders over time.

```{r}
tops <- common_code_counts
#tops$n <- as.character(tops$n)
tops <- tops %>%
  select(underlying, year, n) %>%
  pivot_wider(names_from="year", values_from="n") %>%
  select(underlying, `2003`, `2004`, `2005`, `2006`, `2007`,
         `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`,
         `2015`, `2016`, `2017`, `2018`, `2019`) %>%
  replace(is.na(.), 0)
View(tops)
```

Pivot back
```{r}
get_top_code <- function(idx) {
  o <- tops[idx,2:18] %>%
    pivot_longer(cols=-1) %>%
    rename(year="name") %>%
    rename(n="value") %>%
    select(year, n)
  return(o)
}

o99 <- get_top_code(1)
o26 <- get_top_code(2)
o90 <- get_top_code(3)
o10 <- get_top_code(4)

top_combined <- bind_rows(o99, o26, o90, o10)
```

```{r}
ggplot(o99, aes(x=year, y=n)) + geom_line(group=1)
ggplot(o26, aes(x=year, y=n)) + geom_line(group=1)
ggplot(o90, aes(x=year, y=n)) + geom_line(group=1)
ggplot(o10, aes(x=year, y=n)) + geom_line(group=1)
```

```{r}
o99_main <- mort %>%
  filter(underlying=="O99") %>%
  select(code1, code2, code3, code4, code5, code6, code7, code8, code9, code10)

o99_main
```

```{r}
common_o99 <- mort %>%
  group_by(year) %>%
  add_count(code1)

common_o99_counts <- common_o99 %>%
  group_by(year) %>%
  select(year, code1, n) %>%
  unique() %>%
  slice_max(order_by = n, n = 15) %>%
  arrange(desc(year))

View(common_o99_counts)
```

# Broadening
Because the codes in the dataset lack a certain specificity, it makes sense at this point to look wider instead of deeper. Next up, we are going to explore age.

```{r}
age_groups <- mort %>%
  group_by(year) %>%
  group_by(age) %>%
  select(year, age, underlying, everything()) %>%
  arrange(year, age)

View(age_groups)
```

Add a column for age cohorts, dividing the dataset into periods of five years.

```{r}
mort$age_cohort <- mort$age
mort <- mort %>%
  mutate(age_cohort = case_when(age >= 10 & age <= 15 ~ "10-15",
                                age > 15 & age <= 20 ~ "16-20",
                                age > 20 & age <= 25 ~ "21-25",
                                age > 25 & age <= 30 ~ "26-30",
                                age > 30 & age <= 35 ~ "31-35",
                                age > 35 & age <= 40 ~ "36-40",
                                age > 40 & age <= 45 ~ "41-45",
                                age > 45 & age <= 50 ~ "46-50",
                                age > 50 & age <= 55 ~ "51-55",
                                age > 55 ~ "55+")) %>%
  relocate(age_cohort, .after="age")
```

View cohort size by year.

```{r}
cohort_size <- mort %>%
  group_by(year, age_cohort) %>%
  add_count(age_cohort)

cohort_size_by_year <- cohort_size %>%
  group_by(year) %>%
  select(year, age_cohort, n) %>%
  unique() %>%
  arrange(desc(year)) %>%
  arrange(age_cohort)

print(cohort_size_by_year)
```

Here we see the change in the number of cases of maternal mortality in each cohort. Note steady increases in cohorts especially in 31-40, slight increases in 26-30, and decreases in the 41+ groups.

```{r}
ggplot(cohort_size_by_year, aes(x=year, y=n)) +
  geom_line(group=1) +
  facet_wrap( ~ age_cohort)
```

Get the counts for the top ten codes for each cohort.

```{r}
get_cohort_codes <- function(dataset, cohort, num=10) {
  cohort_codes <- dataset %>%
    filter(age_cohort==cohort) %>%
    select(underlying, age_cohort) %>%
    add_count(underlying) %>%
    unique() %>%
    arrange(desc(n)) %>%
    slice_max(order_by=n, n=num)
  
  return(cohort_codes)
}

cohorts <- mort %>%
  select(age_cohort) %>%
  unique()

cohort_counts <- cohorts[0,]
cohorts <- pull(cohorts, age_cohort)

for (i in 1:length(cohorts)) {
  code_counts <- get_cohort_codes(mort, cohorts[i])
  cohort_counts <- cohort_counts %>%
    bind_rows(code_counts)
}

#View(cohort_counts)
```

To a rough approximation, we can see that there isn't a pronounced change in the kinds of codes that are responsible for the bulk of maternal mortality in different age groups.

```{r}
cohort_counts_charts <- ggplot(cohort_counts, aes(underlying, n)) +
  geom_bar(stat="identity", width=0.5) +
  facet_wrap(~ age_cohort)

View(cohort_counts)
cohort_counts_charts
```

We know from earlier in this analysis that there is a jump in the number of codes. The average number of codes per patient was 1.60 in 2017, and that number immediately jumps to 2.97 in 2018. What will be revealed when we partition the dataset into two along those lines?

```{r}
pre_2018 <- mort %>%
  filter(year<=2017)

post_2018 <- mort %>%
  filter(year>=2018)

# 2018 and 2019 account for almost 20% of the codes assigned, even though these two years are less than 12% of the time frame.
percent_2018 <- (avg_num_codes[16,]$num_codes + avg_num_codes[17,]$num_codes) / sum(avg_num_codes$num_codes) * 100
print(percent_2018)

View(pre_2018)
View(post_2018)
```

Split the dataset on 2018. To see if there is a change in the underlying conditions that are represented.

```{r}
cohorts_pre_2018 <- pre_2018 %>%
  select(age_cohort) %>%
  unique()

cohorts_post_2018 <- post_2018 %>%
  select(age_cohort) %>%
  unique()

cohort_counts_pre_2018 <- cohorts_pre_2018[0,]
cohort_counts_post_2018 <- cohorts_post_2018[0,]

for (i in 1:length(cohorts)) {
  code_counts <- get_cohort_codes(pre_2018, cohorts[i])
  cohort_counts_pre_2018 <- cohort_counts %>%
    bind_rows(code_counts)
}

for (i in 1:length(cohorts)) {
  code_counts <- get_cohort_codes(post_2018, cohorts[i])
  cohort_counts_post_2018 <- cohort_counts %>%
    bind_rows(code_counts)
}

View(cohort_counts_pre_2018)
View(cohort_counts_post_2018)
```

Here we verify that our standouts are much the same. We might have to narrow down on the secondary codes after all.

```{r}
ggplot(cohort_counts_pre_2018, aes(underlying, n)) +
  geom_bar(stat="identity", width=0.5) +
  facet_wrap(~ age_cohort)
```

```{r}
ggplot(cohort_counts_post_2018, aes(underlying, n)) +
  geom_bar(stat="identity", width=0.5) +
  facet_wrap(~ age_cohort)
```

# Additional Codes

As mentioned before, O99, O26, O90, and O10 are our top four codes. It seems that they remain no matter how we break down this dataset, so it behooves us to probe a little farther and see what the additional codes of 2018 and 2019 might be. We will make new tibbles for each of them pre- and post-2018. For the sake of housekeeping, we will also be sure to drop all columns that are completely empty.

```{r}
pre_year_codes <- function(dataset, divide_year, code) {
  pre <- dataset %>%
    filter(year < divide_year) %>%
    filter(underlying == code) %>%
    select(code2:code20) %>%
    select(where(~!all(is.na(.x))))
}
  
post_year_codes <- function(dataset, divide_year, code) {
  post <- dataset %>%
    filter(year >= divide_year) %>%
    filter(underlying == code) %>%
    select(code2:code20) %>%
    select(where(~!all(is.na(.x))))
  
  return(post)
}

o99_pre_2018 <- pre_year_codes(mort, 2018, "O99")
o99_post_2018 <- post_year_codes(mort, 2018, "O99")

o26_pre_2018 <- pre_year_codes(mort, 2018, "O26")
o26_post_2018 <- post_year_codes(mort, 2018, "O26")

o90_pre_2018 <- pre_year_codes(mort, 2018, "O90")
o90_post_2018 <- post_year_codes(mort, 2019, "O90")

o10_pre_2018 <- pre_year_codes(mort, 2018, "O10")
o10_post_2018 <- post_year_codes(mort, 2018, "O10")
```

Get vectors of each non-NA code. Because a high degree of specificity is unnecessary for this aggregation, we are only extracting the main body consisting of three characters. That is, a code like "Y839" will simply become "Y83".

```{r}
library(stringr)

get_secondary_codes <- function(dataset) {
  code_header <- colnames(dataset)
  all_non_na <- c()
  
  for (i in code_header) {
    non_na <- dataset[i] %>%
      drop_na()
    
    all_non_na <- append(all_non_na, non_na)
  }
  
  return(all_non_na)
}

# Using this weird pre- and post- construct because R doesn't allow
# multiple return values :(
a_pre <- get_secondary_codes(o99_pre_2018)
a_post <- get_secondary_codes(o99_post_2018)
b_pre <- get_secondary_codes(o26_pre_2018)
b_post <- get_secondary_codes(o26_post_2018)
c_pre <- get_secondary_codes(o90_pre_2018)
c_post <- get_secondary_codes(o90_post_2018)
d_pre <- get_secondary_codes(o10_pre_2018)
d_post <- get_secondary_codes(o10_post_2018)
```

```{r}
strip_codes_to_base <- function(codes) {
  stripped <- unlist(codes)
  stripped <- str_c(stripped, sep="")
  stripped <- as_vector(stripped)
  stripped <- str_sub(stripped, 1, 3)
}

o99_secondary_pre <- strip_codes_to_base(a_pre)
o99_secondary_post <- strip_codes_to_base(a_post)
o26_secondary_pre <- strip_codes_to_base(b_pre)
o26_secondary_post <- strip_codes_to_base(b_post)
o90_secondary_pre <- strip_codes_to_base(c_pre)
o90_secondary_post <- strip_codes_to_base(c_post)
o10_secondary_pre <- strip_codes_to_base(d_pre)
o10_secondary_post <- strip_codes_to_base(d_post)
```

These are how code counts would be done for the secondary codes.

```{r}
a <- as_tibble(o99_secondary_pre)
a <- a %>%
  select(codes = 1) %>%
  group_by(codes) %>%
  summarise(counts = n()) %>%
  arrange(desc(counts))

a
```

Looking at that table, I can't help but wonder what the breakdown looks like if we only go by the first letter, i.e. look at [the broadest index for what the codes represent](https://www.icd10data.com/ICD10CM/Codes).

```{r}
strip_codes_to_char <- function(codes) {
  stripped <- unlist(codes)
  stripped <- str_c(stripped, sep="")
  stripped <- as_vector(stripped)
  stripped <- str_sub(stripped, 1, 1)
}

get_secondary_counts <- function(codes) {
  counts <- strip_codes_to_char(codes)
  counts <- as_tibble(counts)
  counts <- counts %>%
    select(codes = 1) %>%
    group_by(codes) %>%
    summarise(counts = n()) %>%
    arrange(desc(counts))
  return(counts)
}

o99_secondary_pre_counts <- get_secondary_counts(o99_secondary_pre)
```

Okay, this is an interesting overview of what's happening at the highest level. We should see how this is changing over time for each of the codes.

```{r}
o99_secondary_pre_counts <- get_secondary_counts(o99_secondary_pre)
o99_secondary_post_counts <- get_secondary_counts(o99_secondary_post)
o26_secondary_pre_counts <- get_secondary_counts(o26_secondary_pre)
o26_secondary_post_counts <- get_secondary_counts(o26_secondary_post)
o90_secondary_pre_counts <- get_secondary_counts(o90_secondary_pre)
o90_secondary_post_counts <- get_secondary_counts(o90_secondary_post)
o10_secondary_pre_counts <- get_secondary_counts(o10_secondary_pre)
o10_secondary_post_counts <- get_secondary_counts(o10_secondary_post)
```

There is a lot going on here. Let's just look at all secondary codes starting with I, for circulatory system diseases.

```{r}
cardio <- mort %>%
  filter(str_detect(code2, "I")) %>%
  #select(code2) %>%
  arrange()
```

